---
title: "Examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Two short worked examples. Both run on the package's bundled `cps_sample`
(a 65k-row CPS subsample, 1982–2021) and aim to show the package's two main
modes — descriptive and causal — without doing a full analysis.

* [Example 1: Descriptive within/between decomposition](#e1)
* [Example 2: Causal decomposition — the motherhood penalty](#e2)

```{r setup, message = FALSE, warning = FALSE}
library(ineqx)
data("cps_sample")
```

---

### Example 1: Descriptive within/between decomposition {#e1}

Women in `cps_sample` are grouped into low/medium/high SES (`SES = 1, 2, 3`)
based on household income. We ask: how much earnings inequality lies *within*
each SES tier vs *between* tiers, and how has that changed since 1982?

```{r, message = FALSE}
desc <- ineqx(
  y = "earnweekf", ystat = "CV2",
  group = "SES",
  time = "year", ref = 1982,
  weights = "earnwtf",
  data = cps_sample
)
```

We use `ystat = "CV2"` (squared coefficient of variation, $\sigma^2/\mu^2$)
rather than `"Var"` so that the high-income tier doesn't mechanically dominate
the decomposition. `"VL"` (variance of logs) is also accepted but is generally
not recommended — see the FAQ.

```{r}
print(desc)
```

Within- vs between-group inequality over time:

```{r fig1, fig.width = 8, fig.height = 4, out.width = "85%", dpi = 150, fig.align = "center"}
plot(desc, type = "wibe")
```

Decomposition of the change since 1982 into contributions from changing means
($\Delta_\mu$), changing dispersions ($\Delta_\sigma$), and changing group
shares ($\Delta_\pi$):

```{r fig2, fig.width = 8, fig.height = 4, out.width = "85%", dpi = 150, fig.align = "center"}
plot(desc, type = "decomp")
```

Most of the increase in $CV^2$ since 1982 traces to changes in within-group
dispersion ($\Delta_\sigma$) — earnings inequality has grown *within* SES
tiers, not just *between* them. This is the basic motivation for taking
within-group inequality seriously.

---

### Example 2: Causal decomposition — the motherhood penalty {#e2}

Same dataset, now asking a causal question: how does motherhood affect
earnings inequality among women, and how has that effect changed?

We restrict the sample to women of childbearing age and fit a difference-in-
differences model (treatment = `mother`, post-period = `byear`) with
education-by-decade interactions for both the mean and log-SD of earnings.
GAMLSS does the heavy lifting internally.

```{r, message = FALSE, warning = FALSE}
df <- subset(cps_sample,
             age >= 18 & age <= 49 & !is.na(earnweekf) & earnweekf > 0)

causal <- ineqx(
  y = "earnweekf", ystat = "CV2",
  treat = "mother", post = "byear",
  group = "edu",
  time = "year10", ref = 1980,
  formula_mu    = ~ mother * byear * factor(edu) * factor(year10) + age,
  formula_sigma = ~ mother * byear * factor(edu) * factor(year10),
  weights = "earnwtf",
  se = "delta",
  data = df
)
```

#### Cross-sectional split: het and cov sub-components

The treatment effect on inequality at each time decomposes into
$\tau_B = \text{het}_B + \text{cov}_B$ (between groups) and
$\tau_W = \text{het}_W + \text{cov}_W$ (within groups). See the [Model
structure](model.html) page for the exact formulas. `plot(..., type = "wibe",
stats = ...)` shows all four series at once:

```{r fig3, fig.width = 8, fig.height = 4.5, out.width = "90%", dpi = 150, fig.align = "center", message = FALSE}
plot(causal, type = "wibe",
     stats = c("het_b", "cov_b", "het_w", "cov_w"),
     ci = TRUE)
```

The dominant series here is **`het_W`** — the within-group component driven
by the average treatment effect on log-SD ($\bar\lambda$). It is negative and
large: motherhood compresses within-education earnings dispersion. The
between-group **`cov_B`** term (sorting of mean effects $\beta$ across
education groups) is small and runs in the opposite direction. Note this is
exactly the kind of pattern that mean-based methods miss.

#### Longitudinal: behavioral / compositional / pre-treatment

How has the motherhood effect's contribution to inequality *changed* since
1980? `type = "shapley"` displays the three-channel decomposition with the
range across the six possible orderings:

```{r fig4, fig.width = 8, fig.height = 4.5, out.width = "90%", dpi = 150, fig.align = "center", message = FALSE}
plot(causal, type = "shapley")
```

The **compositional** channel ($\Delta_\pi$) is large and negative — the
share of childbearing-age women who are mothers has dropped substantially
since 1980, and that compositional shift on its own reduces the contribution
of motherhood to overall inequality. Behavioral changes ($\Delta_\beta +
\Delta_\lambda$, how motherhood's per-group effects have changed) are
smaller, and pre-treatment shifts ($\Delta_\mu + \Delta_\sigma$, baseline
inequality changes) contribute a modest negative term.

A more thorough analysis would add controls for race, marital status, age
splines, and family size; would model the pre-trends formally; and would use
bootstrap rather than delta-method SEs. See `?ineqx`, `?boot_config`, and the
[Tutorial](tutorial.html) for the full toolbox.
