---
title: "Examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

I provide two examples on how the ineqx package can be employed to study inequality:

- [Example 1: Tracking changes in within- and between-group inequality over time](#e1)
- [Example 2: Causal decomposition with manual parameters](#e2)

### Tracking changes in within- and between-group inequality over time {#e1}

In this first example, I demonstrate how the ineqx package can be used (1) to split
the inequality in an outcome between groups into within- and between-group components
and (2) to track changes in the within- and between-group components over time.

First, I load a reduced version of the Current Population Survey (CPS) containing a
subsample of all women in CPS with information on their motherhood status and earnings
between 1982 and 2021.

```{r, message = FALSE, warning = FALSE}
library(ineqx)

data("cps_sample")
```

In this dataset, women are grouped into low, medium, and high economic position (SES = 1, 2, 3)
by the total annual income of the household they lead or belong to.

Let us now calculate how much inequality there is within and between those groups:

```{r}
desc <- ineq(
  y = "earnweekf",
  group = "SES",
  time = "year",
  weights = "earnwtf",
  data = cps_sample,
  ref = 1982,
  ystat = "CV2"
)
```

The `ineq()` function takes the information on outcome (y), grouping structure
(group), and time variable (time) to calculate within-group, between-group, and total
inequality in each year. It also decomposes the change over time into contributions
from changing means, dispersions, and composition (Shapley-averaged by default).

```{r}
print(desc)
```

Instead of the variance, we use the squared coefficient of variation as measure of
inequality ($CV^{2}=\sigma^{2}/\mu^{2}$), which equals the variance divided by the
squared mean. This is preferable when decomposing variation into within- and between-group
components to not have the group with the highest income dominate the decomposition.

The resulting within-group and between-group inequality over time:

```{r fig1, fig.width = 8, fig.height = 4, out.width = '75%', dpi=200, fig.align = "center"}
plot(desc)
```

### Causal decomposition with manual parameters {#e2}

In this example, we demonstrate the cross-sectional causal decomposition using the
paper's workers/managers example.

Suppose we have two groups (workers and managers) with equal representation among the
treated, where treatment (e.g., a policy) raises mean earnings by 60 for workers and
100 for managers, and reduces log-SD by 0.1 for workers and 0.2 for managers:

```{r}
params <- ineqx_params(
  data = data.frame(
    group = c("workers", "managers"),
    pi = c(0.5, 0.5),
    mu0 = c(500, 1000),
    sigma0 = c(200, 400),
    beta = c(60, 100),
    lambda = c(-0.1, -0.2)
  )
)

print(params)
```

Now we decompose the treatment effect on inequality:

```{r}
result <- ineqx(params = params, se = "none")
print(result)
```

The between-group effect (delta_B) is positive because managers benefit more from
treatment than workers, which increases between-group inequality. The within-group
effect (delta_W) is negative because treatment compresses the within-group dispersion
(lambda < 0 for both groups).

```{r fig2, fig.width = 6, fig.height = 4, out.width = '60%', dpi=200, fig.align = "center"}
plot(result)
```

If we reverse the treatment effects on means (workers benefit more), the between-group
effect flips sign:

```{r}
params_reversed <- ineqx_params(
  data = data.frame(
    group = c("workers", "managers"),
    pi = c(0.5, 0.5),
    mu0 = c(500, 1000),
    sigma0 = c(200, 400),
    beta = c(100, 60),  # reversed
    lambda = c(-0.1, -0.2)
  )
)

result_reversed <- ineqx(params = params_reversed, se = "none")
print(result_reversed)
```
