---
title: "FAQ"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FAQ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup, echo = FALSE, message = FALSE}
library(ineqx)
```

### Var, CV², or VL — which inequality measure should I use?

`ineqx()` accepts `ystat = "Var"`, `"CV2"`, or (descriptive only) `"VL"`. As a
default, prefer **`"CV2"`** when groups have very different mean levels —
without normalising by the squared mean, the `Var` decomposition is
mechanically dominated by whichever group has the highest mean. Use **`"Var"`**
when levels are comparable across groups or when the dollar-scale
interpretation matters. **Avoid `"VL"`** for substantive decomposition work:
because the log transform is non-linear, the within- and between-group
components of $V_L$ can move in the *opposite direction* from within- and
between-group changes on the income scale. `ineqx()` issues a runtime warning
whenever `"VL"` is used. See the [Model structure](model.html) page §4 for
detail.

### What is the difference between descriptive and causal decomposition?

Both run through the same `ineqx()` function — the dispatch is on whether
`treat` is supplied:

* **Descriptive** (no `treat`): how does total inequality split into within-
  and between-group components, and how have those components shifted over
  time? Calls the Western & Bloome (2009) decomposition.
* **Causal** (with `treat`): given a binary treatment, how does the *treatment
  effect* on inequality split? `ineqx()` fits a GAMLSS model for both the
  mean and log-SD, then decomposes the implied effect on $V$ or $CV^2$ into
  between vs within (and further into heterogeneity vs covariance
  sub-components).

The descriptive case answers *"how unequal is this and where is it growing?"*;
the causal case answers *"how would inequality differ if we removed the
treatment?"* See the [Examples](examples.html) for both.

### Why is `order = "shapley"` the default?

The longitudinal decomposition splits the change in the treatment effect on
inequality into three channels (behavioral, compositional, pre-treatment).
Computing each channel requires a sequential parameter-switching argument,
and the size of any one channel depends on the *order* in which parameters
are switched. With three parameter blocks there are $3! = 6$ orderings.
Shapley averaging takes the unweighted mean across all six and is the
canonical fix for path-dependence in this kind of decomposition. The
`type = "shapley"` plot also displays the *range* across orderings, so you
can see which channels are robust to the choice and which are sensitive.

### When should I use `se = "delta"` vs `se = boot_config()`?

* **`se = "delta"`** (default): closed-form standard errors via the delta
  method, evaluated at the GAMLSS coefficient vcov. Fast, no resampling, and
  available for the integrated estimation mode. Use this unless you have a
  reason not to.
* **`se = boot_config(B = 500, ...)`**: nonparametric bootstrap. Use it when
  (a) you want CIs that don't rely on a normal-approximation, (b) you are
  blending user-supplied params with model params (delta method is not
  supported there), or (c) you want CIs on quantities that aren't covered by
  the closed-form gradients yet. Note that with small `B` (e.g. `B = 10`) the
  percentile CIs will be very noisy — bump to `B = 500+` for production use.
* **`se = "none"`** / `FALSE`: skip SE computation entirely. Useful for
  exploratory work where you just want point estimates fast.

### How do I supply externally estimated parameters?

Use `ineqx_params(data = ...)` to wrap a data.frame with the canonical
columns (`group`, `pi`, `mu0`, `sigma0`, `beta`, `lambda`) into an
`ineqx_params` object, then pass it to `ineqx(..., params = ..., se = "none")`.
This is the right path when you have estimates from a model that doesn't fit
the GAMLSS template — for example a 2SLS specification, or coefficients
copied out of a published table. See `?ineqx_params` for the schema and the
[Tutorial](tutorial.html) for a walk-through.

### What's the difference between `type = "treat.params"` and `type = "outcome.params"`?

Both plot model-implied parameters as a function of group (and time, if
longitudinal). The difference is which parameters:

* **`treat.params`** plots $\beta$ and $\lambda$ — the causal effects of
  treatment on the group mean and log-SD. One panel for each.
* **`outcome.params`** plots $\mu(0), \mu(1) = \mu(0) + \beta$ and $\sigma(0),
  \sigma(1) = \sigma(0) e^{\lambda}$ — the *predicted levels* under control
  and treatment for each group. Useful for showing the underlying potential
  outcomes that drive the decomposition.

Both accept `ci = TRUE` (delta method or bootstrap, whichever is available
on the fitted object) and `style = "line"`/`"point"`.

### My CIs look invisible — what should I check?

A few common causes:

1. **Lowercase `ci`**. The plot argument is `ci = TRUE`, not `CI = TRUE`. R
   is case-sensitive and `CI` silently lands in `...`, so the plot uses the
   default (`ci = FALSE`).
2. **Vertical scale**. If the y-axis spans 0–3 and the SEs are ~0.05, the
   bands are real but only a few percent of the panel — they get easily
   missed in a small figure. Try `style = "point"` to render error *bars*
   instead of ribbons; bars are usually more salient at small SEs.
3. **`se` was set to `"none"`**. If `ineqx()` was called with `se = "none"`
   or `se = FALSE`, no SEs are stored and `ci = TRUE` falls through silently
   (or warns, depending on plot type).
4. **Bootstrap with tiny `B`**. With `B = 10`, percentile bands are dominated
   by noise — they may straddle zero everywhere. Bump to `B ≥ 500`.

### I get an error about `vcov` not being available — what's wrong?

The delta-method SE path needs the GAMLSS coefficient variance-covariance
matrix, which `ineqx()` only stores when both (a) `se = "delta"` was used
*and* (b) the model was fitted via the integrated path (i.e., not blended
with user-supplied params). Two fixes:

* If you used `se = "none"` or fitted via blending: re-run with
  `se = boot_config(B = 500)`. The wibe / treat.params / outcome.params
  plots will fall back to bootstrap percentile CIs automatically.
* If you used `se = "delta"` but the model failed to converge or the vcov
  block is singular: simplify the formulas (`formula_mu` / `formula_sigma`)
  or drop sparsely-populated cells. A degenerate vcov usually points at a
  collinearity issue in the design matrix rather than a problem with `ineqx`.
