---
title: "Getting Started with ineqx"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with ineqx}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  out.width = "80%",
  dpi = 150,
  fig.align = "center"
)
```

This tutorial walks through the three main workflows of the **ineqx** package:

1. [Descriptive decomposition](#descriptive) — split inequality into within- and between-group components
2. [Causal decomposition](#causal) — decompose treatment effects on inequality
3. [External GAMLSS model](#gamlss) — bring your own model, extract parameters, and decompose

The package has three user-facing functions:

| Function | Purpose |
|----------|---------|
| `ineq()` | Descriptive decomposition from raw data |
| `ineqx()` | Causal decomposition (from params or formula + data) |
| `ineqx_params()` | Create parameter object — manually or from a fitted gamlss |

We use the simulated `incdat` dataset that ships with the package. It contains
panel data on individuals across multiple years, three groups
(`group = 1, 2, 3`), and a binary treatment indicator (`x`).

```{r setup}
library(ineqx)

data("incdat")
str(incdat)
```

---

## 1. Descriptive decomposition {#descriptive}

The simplest use case: given an outcome and a grouping variable, how much
inequality is *within* groups versus *between* groups?

```{r desc}
desc <- ineq(
  y     = "inc",
  group = "group",
  time  = "year",
  data  = incdat,
  ref   = 1,
  ystat = "Var"
)

print(desc)
```

When a time variable and reference period are provided, `ineq()`
computes within-group, between-group, and total inequality for each period,
and decomposes the change over time into contributions from changing group
means ($\mu$), dispersions ($\sigma$), and composition ($\pi$), using
Shapley averaging by default.

```{r desc-plot, fig.height = 3.5}
plot(desc)
```

---

## 2. Causal decomposition {#causal}

The causal decomposition answers a different question: *how does a treatment
affect inequality?* It decomposes the treatment's total effect on variance
into:

- **Between-group** (behavioral): does treatment raise means more in some
  groups than others?
- **Within-group** (behavioral): does treatment compress or expand dispersion
  within groups?

### Integrated fitting (Mode B)

The simplest approach: pass formulas and data directly to `ineqx()`, which
fits a GAMLSS model internally:

```{r causal-modeb, message = FALSE}
result <- ineqx(
  formula_mu    = inc ~ x * factor(group),
  formula_sigma = ~ x * factor(group),
  data  = incdat,
  treat = "x",
  group = "group",
  se    = "none"
)

print(result)
```

```{r causal-cross-plot, fig.height = 3.5}
plot(result)
```

### Pre-extracted parameters (Mode A)

For more control, create an `ineqx_params` object first and pass it to
`ineqx()`:

```{r causal-modea, message = FALSE}
# Extract params from a gamlss model
params <- ineqx_params(
  model = gamlss::gamlss(
    inc ~ x * factor(group) + factor(t),
    sigma.formula = ~ x * factor(group) + factor(t),
    data = incdat, family = gamlss.dist::NO(), trace = FALSE
  ),
  data  = incdat,
  treat = "x",
  group = "group",
  time  = "t",
  ref   = 0,
  vcov  = FALSE
)

print(params)
```

### Longitudinal causal decomposition

With longitudinal parameters, `ineqx()` automatically performs the
longitudinal decomposition:

```{r causal-longit}
result_longit <- ineqx(
  params = params,
  order  = c("behavioral", "compositional", "pretreatment"),
  se     = "none"
)

print(result_longit)
```

```{r causal-longit-plot, fig.height = 3.5}
plot(result_longit)
```

### Shapley values

Since the longitudinal decomposition depends on the ordering, you can
average across all 6 orderings with `order = "shapley"`:

```{r shapley}
result_shapley <- ineqx(params = params, order = "shapley", se = "none")
print(result_shapley)
```

```{r shapley-plot, fig.height = 3.5}
plot(result_shapley)
```

---

## 3. Using an external GAMLSS model {#gamlss}

If you want full control over the model specification (e.g., different
distributions, random effects, custom controls), fit a `gamlss` model
yourself and pass it to `ineqx_params()`:

```{r external-gamlss, message = FALSE}
library(gamlss)

# Fit your own GAMLSS
my_model <- gamlss(
  formula       = inc ~ x * factor(group) + factor(t),
  sigma.formula = ~ x * factor(group) + factor(t),
  data          = incdat,
  family        = NO(),
  trace         = FALSE
)

# Extract parameters
params_ext <- ineqx_params(
  model = my_model,
  data  = incdat,
  treat = "x",
  group = "group",
  vcov  = TRUE
)

print(params_ext)

# Decompose
ineqx(params = params_ext)
```

### Manual parameter specification

You can also construct `ineqx_params` by hand — useful for simulation
studies, sensitivity analyses, or replicating results from other software:

```{r manual-params}
params_manual <- ineqx_params(
  data = data.frame(
    group  = c("low", "mid", "high"),
    pi     = c(0.4, 0.35, 0.25),
    mu0    = c(300, 600, 1200),
    sigma0 = c(100, 200, 500),
    beta   = c(50, 80, 120),
    lambda = c(-0.05, -0.10, -0.15)
  ),
  ystat = "Var"
)

ineqx(params = params_manual, se = "none")
```

---

## Summary

| Workflow | Functions | Input |
|----------|-----------|-------|
| Descriptive | `ineq()` | Raw data (y, group, time) |
| Causal (integrated) | `ineqx(formula_mu=..., data=...)` | Raw data + treatment |
| Causal (external model) | `ineqx_params(model=...)` + `ineqx(params=...)` | Fitted gamlss |
| Causal (manual) | `ineqx_params(data=...)` + `ineqx(params=...)` | Parameter data.frame |

For details on the methodology, see `vignette("model")`. For more examples, see `vignette("examples")`.
