---
title: "Getting Started with ineqx"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with ineqx}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
rm(list=ls())
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  out.width = "80%",
  dpi = 150,
  fig.align = "center"
)
```

This tutorial walks through the three main workflows of the **ineqx** package:

1. [Descriptive decomposition](#descriptive) — split inequality into within- and between-group components
2. [Causal decomposition](#causal) — decompose treatment effects on inequality
3. [External GAMLSS model](#gamlss) — bring your own model, extract parameters, and decompose

The package has two user-facing functions:

| Function | Purpose |
|----------|---------|
| `ineqx()` | Variance decomposition — descriptive (no `treat`) or causal (with `treat` or `params`) |
| `ineqx_params()` | Create parameter object — manually or from a fitted gamlss |

We use the simulated `incdat` dataset that ships with the package. It contains
panel data on individuals across multiple years, three groups
(`group = 1, 2, 3`), and a binary treatment indicator (`x`).

```{r setup}
devtools::load_all()

data("incdat")
str(incdat)
```

---

## 1. Descriptive decomposition {#descriptive}

The simplest use case: given an outcome and a grouping variable, how much
inequality is *within* groups versus *between* groups? When no `treat`
argument is provided, `ineqx()` performs a descriptive decomposition.

```{r desc}
desc <- ineqx(
  y     = "inc",
  group = "group",
  time  = "year",
  ref   = 1,
  ystat = "Var",
  data  = incdat
)

print(desc)
```

When a time variable and reference period are provided, `ineqx()`
computes within-group, between-group, and total inequality for each period,
and decomposes the change over time into contributions from changing group
means ($\mu$), dispersions ($\sigma$), and composition ($\pi$), using
Shapley averaging by default.

```{r desc-plot-deltas, fig.height = 3.5}
plot(desc)
```

The default plot type is `"decomp"`, which shows how inequality changed relative
to the reference period and decomposes that change into $\mu$, $\sigma$, and
$\pi$ contributions.

### Within/between levels

The `type = "wibe"` plot shows the raw within-group, between-group, and total
inequality at each time point:

```{r desc-wibe, fig.height = 3.5}
plot(desc, type = "wibe")
```

### Group-level parameters

The `type = "params"` plot shows group-level means and standard deviations over
time. Pass `ci = TRUE` or `ci = "delta"` to add confidence intervals:

```{r desc-params, fig.height = 4}
plot(desc, type = "params")
```

### Inequality statistics

The `type = "ineq"` plot computes and displays overall inequality statistics
over time. You can choose from `"V"` (variance), `"VL"` (log-variance),
`"CV2"`, `"Gini"`, and `"Theil"`, or pass custom functions:

```{r desc-ineq, fig.height = 3.5}
plot(desc, type = "ineq", stats = c("V", "Gini"))
```

The `type = "ineq.group"` variant shows per-group inequality statistics:

```{r desc-ineq-group, fig.height = 3.5}
plot(desc, type = "ineq.group", stats = c("V"))
```

### Counterfactual reference

You can decompose inequality relative to a counterfactual baseline by
passing a descriptive params object via `params`. Create it with
`ineqx_params()` using columns `group`, `pi`, `mu`, `sigma` — this
parallels the causal path where `params` supplies the reference state.

For example, to decompose the observed inequality into contributions from
group means, dispersions, and composition relative to a world of perfect
equality (all parameters zero, equal group shares):

```{r desc-ref}
ref <- ineqx_params(
  data = data.frame(
    group = 1:3,
    pi    = 1/3,
    mu    = 0,
    sigma = 0
  )
)

desc_ref <- ineqx(
  y     = "inc",
  ystat = "Var",
  group = "group",
  time  = "year",
  params = ref,
  data  = incdat
)

print(desc_ref)
```

The deltas show how much each parameter contributes to the level of
inequality at each time period: $\mu$ drives between-group inequality,
$\sigma$ drives within-group inequality, and $\pi$ captures the effect
of unequal group sizes.

---

## 2. Causal decomposition {#causal}

The causal decomposition answers a different question: *how does a treatment
affect inequality?* It decomposes the treatment's total effect on variance
into:

- **Between-group** (behavioral): does treatment raise means more in some
  groups than others?
- **Within-group** (behavioral): does treatment compress or expand dispersion
  within groups?

### Integrated estimation

The simplest approach: pass formulas and data directly to `ineqx()`, which
fits a GAMLSS model internally. Note that `formula_mu` is one-sided — the
outcome is specified via the `y` argument:

```{r causal-integrated, message = FALSE}
result <- ineqx(
  y             = "inc",
  treat         = "x",
  group         = "group",
  formula_mu    = ~ x * factor(group),
  formula_sigma = ~ x * factor(group),
  se            = "delta",
  data          = incdat
)

print(result)
```

The default `type = "wibe"` shows the within- and between-group treatment
effects as a bar chart. The `show` argument controls which sub-components
to display (default: `c("tau", "cov")`):

```{r causal-cross-plot, fig.height = 3.5}
plot(result)
```

### Group-level contributions

The `type = "wibe.group"` plot shows each group's contribution to the
within-group and between-group treatment effects:

```{r causal-cross-wibe-group, fig.height = 3.5}
plot(result, type = "wibe.group")
```

### Treatment effect parameters

The `type = "treat.params"` plot shows the estimated treatment effect
parameters ($\beta$ and $\lambda$) by group:

```{r causal-cross-params, fig.height = 3.5}
plot(result, type = "treat.params")
```

### Predicted outcomes

The `type = "outcome.params"` plot compares predicted group-level means and
standard deviations under control versus treatment:

```{r causal-cross-outcome, fig.height = 3.5}
plot(result, type = "outcome.params")
```

### Treatment effect distributions

The `type = "treat"` plot overlays the predicted treatment effect
distributions for each group. Use `type = "outcome"` to see the full
predicted outcome distributions under control vs treatment:

```{r dist-cross, fig.height = 5}
plot(result, type = "treat")
```

```{r dist-cross-outcome, fig.height = 5}
plot(result, type = "outcome")
```

### Externally estimated model

For more control, create an `ineqx_params` object first and pass it to
`ineqx()`:

```{r causal-external, message = FALSE}
# Extract params from a gamlss model
params <- ineqx_params(
  model = gamlss::gamlss(
    inc ~ x * factor(group) + factor(year),
    sigma.formula = ~ x * factor(group) + factor(year),
    data = incdat, family = gamlss.dist::NO()
  ),
  data  = incdat,
  treat = "x",
  group = "group",
  time  = "year",
  vcov  = TRUE
)

print(params)
```

### Longitudinal causal decomposition

With longitudinal parameters, `ineqx()` automatically performs the
longitudinal decomposition:

```{r causal-longit}
result_longit <- ineqx(
  "inc",
  group  = "group",
  ref    = 1,
  order  = c("behavioral", "compositional", "pretreatment"),
  params = params,
  se     = "delta",
  data   = incdat
)

print(result_longit)
```

The default `type = "decomp"` shows the three-component decomposition
(behavioral, compositional, pre-treatment) of changes relative to the
reference:

```{r causal-longit-plot, fig.height = 3.5}
plot(result_longit)
```

### Six sub-component decomposition

Setting `show` to individual delta components displays the six
sub-components ($\Delta_\beta$, $\Delta_\lambda$, $\Delta_{\pi,b}$,
$\Delta_{\pi,w}$, $\Delta_\mu$, $\Delta_\sigma$):

```{r causal-longit-six, fig.height = 3.5}
plot(result_longit, type = "decomp",
     show = c("delta_beta", "delta_lambda", "delta_pi_b",
              "delta_pi_w", "delta_mu", "delta_sigma"))
```

### Within- and between-group treatment effects

The `type = "wibe"` plot shows the *levels* of the cross-sectional treatment
effects ($\tau_B$, $\tau_W$) at each time point, rather than changes from
the reference. The `show` argument controls which sub-components to display
(default: `c("tau_b", "tau_w")`; also available: `"het_b"`, `"cov_b"`,
`"het_w"`, `"cov_w"`):

```{r causal-wibe, fig.height = 3.5}
plot(result_longit, type = "wibe")
```

### Shapley values

Since the longitudinal decomposition depends on the ordering, you can
average across all 6 orderings with `order = "shapley"`:

```{r shapley}
result_shapley <- ineqx("inc", group = "group", ref = 1,
                         params = params, order = "shapley",
                         se = "none", data = incdat)
print(result_shapley)
```

```{r shapley-plot, fig.height = 3.5}
plot(result_shapley)
```

The `type = "shapley"` plot shows the Shapley-averaged values with ribbons
indicating the range across orderings — useful for assessing ordering
dependence:

```{r shapley-ranges, fig.height = 3.5}
plot(result_shapley, type = "shapley")
```

### Treatment effect parameters over time

The `type = "treat.params"` plot shows the estimated treatment effect
parameters ($\beta$ and $\lambda$) by group over time:

```{r params-longit, fig.height = 3.5}
plot(result_shapley, type = "treat.params")
```

### Predicted outcomes over time

The `type = "outcome.params"` plot compares predicted group-level means and
standard deviations under control versus treatment over time:

```{r outcome-longit, fig.height = 3.5}
plot(result_shapley, type = "outcome.params")
```

### Predicted distributions

The `type = "treat"` plot shows predicted treatment effect distributions.
When `time` is specified, it shows per-group distributions at that time
point. When `time = NULL` (default), it shows pi-weighted marginal
distributions across all time points. Use `type = "outcome"` for predicted
outcome distributions:

```{r dist-longit, fig.height = 5}
plot(result_shapley, type = "treat", time = 3)
```

```{r dist-longit-marginal, fig.height = 5}
plot(result_shapley, type = "treat")
```

### Counterfactual reference (causal)

Just as the descriptive decomposition supports a counterfactual baseline,
the causal decomposition can blend user-provided counterfactual parameters
with model-estimated parameters from data. This answers questions like:
*"How did treatment effects on inequality evolve relative to a hypothetical
baseline?"*

Provide an `ineqx_params` object (with causal columns `mu0`, `sigma0`,
`beta`, `lambda`) together with `treat`, `formula_mu`, and `formula_sigma`.
The function fits a GAMLSS model on the data, then blends the
model-estimated parameters with your counterfactual scenarios.

You can specify counterfactual parameters at multiple time points. Periods
covered by your params override the model estimates, while remaining periods
are estimated from data. Here we place two counterfactual scenarios — one at
the beginning (time 0: no treatment effect) and one in the middle (time 3:
moderate, uniform treatment effect) — with data-estimated periods in between:

```{r causal-blend, message = FALSE}
# Two counterfactual scenarios at time 0 and time 3
cf_ref <- ineqx_params(
  data = data.frame(
    group  = rep(1:3, 2),
    time   = rep(c(0, 3), each = 3),
    pi     = 1/3,
    mu0    = 10,
    sigma0 = 1,
    # time 0: no treatment effect
    # time 3: moderate uniform treatment effect
    beta   = c(0, 0, 0, 5, 5, 5),
    lambda = c(0, 0, 0, 1, 1, 1)
  )
)

result_blend <- ineqx(
  y             = "inc",
  treat         = "x",
  group         = "group",
  time          = "year",
  params        = cf_ref,
  ref           = 0,
  formula_mu    = ~ x * factor(group) + factor(year),
  formula_sigma = ~ x * factor(group) + factor(year),
  se            = "none",
  data          = incdat
)

print(result_blend)
```

Time 0 and time 3 use the counterfactual parameters; times 1, 2, 4, and 5
are estimated from data via GAMLSS. The longitudinal decomposition shows
how treatment effects on inequality evolved across all periods relative to
the reference (time 0).

Note: delta method SEs are not available for blended parameters. Use
`boot_config()` for bootstrap SEs if needed.

### Comparing scenarios

The `compare()` function stacks multiple ineqx results for side-by-side
comparison. This is useful for contrasting an observed decomposition with
one or more counterfactual baselines:

```{r compare, fig.height = 3.5}
comp <- compare(Observed = result_shapley, Counterfactual = result_blend)
print(comp)
```

The default `type = "decomp"` plot uses color for the decomposition component
and linetype for the scenario:

```{r compare-decomp, fig.height = 3.5}
plot(comp, type = "decomp")
```

Use `show` to display individual sub-components:

```{r compare-decomp-six, fig.height = 3.5}
plot(comp, type = "decomp",
     show = c("delta_beta", "delta_lambda", "delta_pi_b",
              "delta_pi_w", "delta_mu", "delta_sigma"))
```

The `type = "wibe"` plot compares within/between treatment effect levels
across scenarios:

```{r compare-wibe, fig.height = 3.5}
plot(comp, type = "wibe")
```

The underlying stacked data is accessible via `comp$data`.

---

## 3. Using an external GAMLSS model {#gamlss}

If you want full control over the model specification (e.g., different
distributions, random effects, custom controls), fit a `gamlss` model
yourself and pass it to `ineqx_params()`:

```{r external-gamlss, message = FALSE}
library(gamlss)

# Fit your own GAMLSS
my_model <- gamlss(
  formula       = inc ~ x * factor(group) + factor(t),
  sigma.formula = ~ x * factor(group) + factor(t),
  data          = incdat,
  family        = NO(),
  trace         = FALSE
)

# Extract parameters
params_ext <- ineqx_params(
  model = my_model,
  data  = incdat,
  treat = "x",
  group = "group",
  vcov  = TRUE
)

print(params_ext)

# Decompose
ineqx(y="inc", group = "group", params = params_ext, data = incdat)
```

### Manual parameter specification

You can also construct `ineqx_params` by hand — useful for simulation
studies, sensitivity analyses, or replicating results from other software:

```{r manual-params}
params_manual <- ineqx_params(
  data = data.frame(
    group  = c("low", "mid", "high"),
    pi     = c(0.4, 0.35, 0.25),
    mu0    = c(300, 600, 1200),
    sigma0 = c(100, 200, 500),
    beta   = c(50, 80, 120),
    lambda = c(-0.05, -0.10, -0.15)
  ),
  ystat = "Var"
)

ineqx(y="inc", group = "group", params = params_manual,
      se = "none", data = incdat)
```

---

## Summary

| Workflow | Call | Input |
|----------|------|-------|
| Descriptive | `ineqx(y, group, data=...)` | Raw data |
| Descriptive (counterfactual ref) | `ineqx_params(data=...)` then `ineqx(y, group, params=..., data=...)` | Reference params + raw data |
| Causal (integrated) | `ineqx(y, treat, group, formula_mu=..., data=...)` | Raw data + treatment |
| Causal (counterfactual ref) | `ineqx_params(data=...)` then `ineqx(y, treat, group, params=..., formula_mu=..., data=...)` | Reference params + raw data + treatment |
| Causal (external model) | `ineqx_params(model=...)` then `ineqx(y, group, params=..., data=...)` | Fitted gamlss |
| Causal (manual) | `ineqx_params(data=...)` then `ineqx(y, group, params=..., data=...)` | Parameter data.frame |

For details on the methodology, see `vignette("model")`. For more examples, see `vignette("examples")`.
